Exploring the Role of Feedback in Machine Learning Jupyter Notebooks

### 2024-06-10 report/emse24.tex
feat(report): emse24 updates

\subsection{Jupyter Notebooks}

% NOTE use background section from pimentel2019large-scale

@@

@@

@@

\begin{figure}
\centering
\includegraphics[width=\linewidth]{data-collection.pdf}
\label{fig:data-collection}
\caption{Overview of data collection methodology used in this study}
\end{figure}

@@

\section{Threats to Validity}\label{sec:threats}
TODO

@@

\subsection{Feedback in Software Engineering and Machine Learning}

% the role of early adoption of feedback when developing software, agile
% similar concept exists in ML: crisp-dm, mldl, sculley, amershi, sambasivan
% feedback from one part of the pipeline feeds into another,

@@

% start with a very quick review of the workflow: REPL style of development; break code into smaller chunks; faster feedack; iterative style of development
% link to the existing work on notebooks from chi: design, benefits and challenges
% then go into the SE side of software written in notebooks: code quality
% adoption of notebooks by ML community

% start with a quick introduction to the anatomy of a computational notebook: code is broken down into smaller chunks represented as cells, cells can contain code (which are executable and immediately display the result of the execution), or markdown cells used for the purpose of documenting or creating a narrative for collaborative work.

% the move on to the benefits of this style of coding: the ML pipeline contains several components, which can be broken down into smaller, more manageable chunks. These can be individually developed incorporating the rapid feedback provided from the REPL style of coding.

% then summarise the pain-points of working in notebooks: use chattopadhyay2020whats
Computational notebooks have been ubiquitously adopted by the ML community for developing ML enabled systems. Although originally intended to promote reproducible software, computational notebooks are far from being reproducible due to lack of software engineering best practices such as separation of concern, testing, dependenvcy management and versioning~\cite{pimentel2019large-scale,wang2020better,chattopadhyay2020whats}.

@@

Testing ML enabled software systems poses more challenges compared to
their traditional counter part. While traditional software systems
mature through change in their codebase, ML enabled systems
additionally experience change in the training data and the machine
learning
model~\cite{sculley2015hidden,amershi2019software,sambasivan2021everyone}.
With ML being adopted into safety-critical domains that can affect
human lives, ensuring that ML systems are correct, robust to data
perturbations and not biased by race, colour or gender is of paramount
importance. Existing scientific literature on ML testing broadly
focuses on two aspects. First, on functional properties such as
correctness and robustness of the model towards unseen data. And
second, on non-functional properties such as fairness,
interpretability and
privacy~\cite{zhang2020machine,mehrabi2021survey,chen2022fairness}.

To test the correctness of ML models, several improvements over
existing test adequacy metrics have been proposed. Tools such as
\textit{DeepExplore} and \textit{DeepGauge} propose new test adequacy
metrics such as neuron coverage adapted for ML enabled software
systems~\cite{pei2017deepexplore, ma2018deepgauge,
  gerasimou2020importance}. Formal verification methods have also been
proposed that try to provide formal guarantee of robustness against
adversarial examples. Such methods however are only feasible for
statistical ML models and become computationally expensive for more
complex models such as deep neural networks~\cite{zhu2021deepmemory,
  baluta2021scalable}. Several works have been conducted on generating
and detecting adversarial inputs for ML models. Data augmentation
techniques based on fuzzing, search based software testing and
mutation testing have been proposed to generate adversarial examples
that can be used during model training to improve its
robustness~\cite{braiek2019deepevolution, gao2020fuzz, wang2021robot,
  zhang2020white}. It is however not possible to include all
variations of adversarial examples into the training data. Thus,
methods have been proposed to detect adversarial inputs during
runtime~\cite{xiao2021self, wang2020dissector, wang2019adversarial,
  berend2020cats}.

@@

As ML augment software systems in safety-critical domains, emphasis
has been put into explainability of ML models. ML models and the
underlying data is complex and multi-dimensional. To combat the
``curse of dimensionality'', visual analytics has been widely adopted
by the ML community to understand the data and the internal workings
of ML models~\cite{yuan2021survey,hohman2019visual,wexler2020if}.

@@

\section{Future Work}

@@

\section{Conclusion}

>>>>

\subsection{Literate Programming, Computational Notebooks and Project Jupyter}

% NOTE pimentel2019large-scale:
% NOTE kery2018story
% NOTE rule2018exploration

@@

\begin{figure}
  \centering
  \includegraphics[width=0.75\linewidth]{nb.pdf}
  \label{fig:nb}
  \caption{Example Jupyter notebook with code and markdown cells, adapted from~\cite[Figure~1]{pimentel2019large-scale}.}
\end{figure}

Literate programming is a paradigm introduced by Donald Knuth that emphasizes writing code that is understandable by humans. In his seminal work, Knuth adovocates combining source code with documentation in a single, cohesive narrative~\cite{knuth1984literate}. Computational notebooks are inspired by this paradigm, allowing for the integration of executable code, scientific equations, visualizations, and explanatory text within a single document. This format supports a literate approach to programming, making it easier to share and understand complex analyses and workflows. Project Jupyter is a prominent example of this approach, offering a suite of open-source tools for interactive computing. Jupyter Notebooks, in particular, have become a popular tool in the ML community since they facilitate open and reproducible science, education, and data analysis.

% front-end which parses the JSON
% there is also the kernel which is a process running in the background, the frontend communicates with the kernel; send input (code cell) and get output (result of execution)

Figure~\ref{fig:nb} shows an Jupyter notebook with markdown and code cells. The data comprised within Jupyter Notebooks is stored using the JSON schema specified by \emph{nbformat}~\footnote{https://nbformat.readthedocs.io/en/latest/index.html}, and thus can be parsed programmatically. A notebook consists of \emph{cells} which can be of three types---``markdown'', ``code'' and ``raw''. All cells of a notebook are stored as a list under the top-level \lstinline[language={}]$cells$ key in the JSON representation. Within this list, each cell is represented as a JSON object. The contents of the cell are stored as a list of strings inside the \lstinline[language={}]$source$ key.

@@

An Abstract Syntax Tree (AST) is a tree representation of the abstract syntactic structure of source code. Each node in the tree represents a construct occurring in the source code. ASTs are widely used in compilers and code analysis tools to understand and manipulate code programmatically~\cite{wikipedia2024abstract}. The Python \emph{ast} module enables the parsing, manipulation, and analysis of Python source code by converting it into an AST. This module allows for the examination and modification of code structure, making it valuable for tasks such as code analysis, optimization, and transformation. The ast module can parse source code strings into AST objects, traverse and modify the tree, and compile it back into executable code, providing a powerful tool for dynamic code manipulation~\cite{pythonast}.

In Python, \lstinline{assert} statements are represented in the AST by the \lstinline{Assert} class. The \lstinline{Assert} node includes the \lstinline{test} attribute---which contains the condition being asserted---and the \lstinline{msg} attribute---which holds an optional message displayed if the assertion fails. \lstinline{print} statements are treated as a call to the built-in \lstinline{print} function. Function calls are represented in the AST by the \lstinline{Call} class. The \lstinline{Call} node includes \lstinline{func}, representing the \lstinline{print} function, and \lstinline{args} which lists the arguments passed.

@@

\begin{figure}
  \centering
  \includegraphics[width=0.75\linewidth]{data-collection.pdf}
  \label{fig:data-collection}
  \caption{Overview of data collection methodology used in this study}
\end{figure}

@@

\section{Threats to Validity}\label{sec:threats}

The empirical analysis presented in this study uses Jupyter notebooks mined from Github. To the best of our knowledge,~\emph{Pimentel et al} and~\emph{Quaranta et al} are the only two prior studies which provide a public dataset of Jupyter Notebooks. We faced several technical challenges while trying to access the notebooks provided by these replication packages and thus decided to mine our own data from Github. However, 24\% of the notebooks had to be excluded from the study since they were not related to ML or DS. We wish to extend our analysis to notebooks collected from Kaggle since it is predominantly targeted towards the ML and DS community.

In Section~\ref{sec:data-collect}, we intentionally keep the search criteria on Github broad by only checking for the appropriate file type and the presence of the ``assert'' keyword. Besides being used to define tests in the Python programming language, ``assert'' also appears in several external python testing libraries (for instance nose~\footnote{http://readthedocs.org/docs/nose/}, unittest~\footnote{https://docs.python.org/3/library/unittest.html} and numpy~\footnote{https://numpy.org/doc/stable/reference/routines.testing.html}). Our approach captures a rich variety of assertion statements while preventing the need to craft a custom search query based on an exhaustive list of keywords that appear in python testing libraries.

@@

@@

@@

% Testing ML enabled software systems poses more challenges compared to their traditional counter part. While traditional software systems mature through change in their codebase, ML enabled systems additionally experience change in the training data and the machine learning model~\cite{sculley2015hidden,amershi2019software,sambasivan2021everyone}. With ML being adopted into safety-critical domains that can affect human lives, ensuring that ML systems are correct, robust to data perturbations and not biased by race, colour or gender is of paramount importance.

Existing scientific literature on ML testing broadly focuses on two aspects. First, on functional properties such as correctness and robustness of the model towards unseen data. And second, on non-functional properties such as fairness, interpretability and privacy~\cite{zhang2020machine,mehrabi2021survey,chen2022fairness}.

To test the correctness of ML models, several improvements over existing test adequacy metrics have been proposed. Tools such as \textit{DeepXplore} and \textit{DeepGauge} propose new test adequacy metrics such as neuron coverage adapted for ML enabled software systems~\cite{pei2017deepxplore, ma2018deepgauge, gerasimou2020importance-driven}. Formal verification methods have also been proposed that try to provide formal guarantee of robustness against adversarial examples. Such methods however are only feasible for statistical ML models and become computationally expensive for more complex models such as deep neural networks~\cite{zhu2021deepmemory, baluta2021scalable}. Several works have been conducted on generating and detecting adversarial inputs for ML models. Data augmentation techniques based on fuzzing, search based software testing and mutation testing have been proposed to generate adversarial examples that can be used during model training to improve its robustness~\cite{braiek2019deepevolution, gao2020fuzz, wang2021robot, zhang2020white-box}. It is however not possible to include all variations of adversarial examples into the training data. Thus, methods have been proposed to detect adversarial inputs during runtime~\cite{xiao2021self-checking, wang2020dissector, wang2019adversarial, berend2020cats}.

@@

As ML augment software systems in safety-critical domains, emphasis has been put into explainability of ML models. ML models and the underlying data is complex and multi-dimensional. To combat the ``curse of dimensionality'', visual analytics has been widely adopted by the ML community to understand the data and the internal workings of ML models~\cite{yuan2021survey,hohman2019visual,wexler2019what-if}.

@@

\section{Conclusion and Future Work}

@@

This study empirically supports the efficacy of visualisations for testing various properties of machine learning systems. Visualisations enhance the ability of ML practitioners to interpret complex information through visual cues. However, decisions based on such visualisations demand manual verification in each development iteration. Analytical assertions derived from ML visualisations can reduce this manual effort by encapsulating key insights, and ensuring continuity in interpretation across different development stages and project collaborators.

Our results indicate that crafting analytical assertions from ML visualisations, is an emerging testing technique. Majority of the assertions however, are found to be primitive and unable to convey all information discerned from their visual counter-parts. This study finds a gap in the current software testing methods to address the unique challenges of ML. And emphasises the need for automated tools to bridge visual assessments and analytical assertions.

There is a clear need for tools designed to aid ML practitioners when testing ML systems. These tools should provide assistance in identifying the most suitable visualisations for testing specific properties of ML systems. Customized solutions that streamline the process of connecting visual insights to analytical assertions would significantly benefit practitioners in this domain.

The manual translation of insights derived from visualisations into analytical assertions presents a substantial challenge. Future work should focus on the development of automated tools that can assist AI practitioners in this translation process. Generating assertions from visual information automatically, will promote adoption of assertions in ML notebooks. And bring the code developed within notebooks, closer to a production-ready codebase.

% TODO remove this is no space left
Extending ML testing to dynamic real-world scenarios should be explored further. This entails developing techniques and frameworks that allow assertions to adapt to evolving data distributions, changing model requirements, and continuous system updates. Research in this area will contribute to the development of ML systems that remain reliable and relevant throughout their operational lifecycle.
